### 2024-07-05 thesis/ms-fletcher-porter.tex
Reworked introduction
A brief history of control tooling

In order to study mechatronics systems, it is necessary to design and develop physical research test beds. This is especially the case in mechatronic systems with fluid components, as their high-fidelity modeling is challenging and computationally expensive. When doing research with hardware systems, before a single research question can be asked of the hardware, researchers must assemble mechanical components, hook up power electronics, and route hoses for the hydraulic and pneumatic systems. Then the most complex of them all, the control system, must be added. Its tangle of spindly wires reaches all throughout the machine, meeting at the computer, the single most complex component of any fluid power system. The requirements placed on it are immense. In high-power systems, safety invariants must be held up perfectly lest disaster strike, and close monitoring is necessary due to the low stiffness of fluid systems compared to mechanical ones.

This paper improves the process of commissioning control systems for fluid power systems by reasoning about how to abstract behavior and components in control systems into layers, synthesizing a novel {\it mechatronic tech stack\/} theory. It will demonstrate in three case studies how this theory can be applied in fluid power systems. The tools used in these cases will be categorized into layers of the mechatronic tech stack.  The analysis shows that, while different, these cases all fit into the same structure.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/osi_model}
  \caption{The OSI model understands network communication to be a layer cake of abstractions over the physical communication media.  An individual tool may implement the behviours and interfaces of of one or multiple layers.  The layering enables design freedom in how a network is implemented.  Communication over HTTP usually goes over a TCP transport through an IP network and to other destinations over WiFi or Ethernet, but Lindgren et al.\ show that it's possible to send IP signals over CAN~\cite{Lindgren2008}, and Ethernet over EtherCAT is possible as well~\cite[§1.9.3]{EtherCATFieldbus}.  High level protocols can also be replaced to obtain a lower level of abstraction for performance or design flexibility, as was done for the video game \emph{DOOM} in 1993, as well as for contemporary game engines like Godot or Unity~\cite{DOOM1993,GodotMultiplayer,UnityTransport}.}\label{fig:osi_model}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/printf_syntax}
  \cprotect\caption{The above printf example with syntax labeled according to Kernighan {\&} Ritchie C~\cite{Kernighan1978}.  The whole line is a postfix expression and the \verb|printf| token is also a postfix expression on its own.  The string between parentheses is an argument expression list, and the strings between commas are a variety of syntactic expressions that enable writing non-trivial code, terminating as a primary expression, and then here has a string literal and an identifier.}\label{fig:printf_syntax}
\end{figure}

The semantics of \verb|printf| are the exprected behavior of the function, which can be gleaned from the documentation. It will say that \verb|printf| means to convert and write output to the console under the control of a format string~\cite{Kernighan1978}.  It will also say that the format string is the first expression in th argument expression list and that it is of type \verb|char*|.  So \verb|"I have %u cows\n"| is that format string, and the documentation will state that the second expression in the argument expression list, here \verb|num_cows| will be formatted as a string and written to the output where conversion character, here \verb|%u| is.  The logic of how the conversion characters and string formatting work are based on the format string's own morphology, syntax, and semantics, the study of which is left as an exercise to the reader.

@@

Syntax is the structure of an idea. It is a finer comb to discriminate layers. Software generally is built as a linear sequence of instructions with branches and loops. This describes the control engineer's perspective on their own control program, but supporting software like libraries are more abstract. They're better thought of as a dependency. A control program may depend on libraries A, B, and C, and those in turn depend on libraries D, E, F, and G. These \textit{middlewares}, named so as they exist in the middle between hte control software and the computer, are better understood as a network of the programs themselves and the dependency connections between them. For hardware, its structure is its assemblage. The computer components are connected to the motherboard, sensors and transducers are connected to a CAN bus, the temperature sensors message over I2C, etc. Like middleware, hardware is a network of components and the connections between them, but the type of connection is also important. A system with components on a CAN bus is going to be different than one with a USB bus, and different still from a system where all components communicate over a serial connection.

@@

This is enough to propose a mechatronic tech stack. The mechatronic stack as outlined sits on top of signaling with sensors and transducers. Those are interfaced by IO systems that connect to computer systems which run middleware upon which a control program depends. The design is so that functionality in the control layer should be preserved if components in the lower layers are changed, perhaps wih some implementation effort, as was the case for IP over CAN.  This is illustrated as a stack diagram in figure~\ref{fig:mechatronic_tech_stack}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/mechatronic_tech_stack}
  \caption{The mechatronic tech stack places high-level business logic at the top and low-level interfacing at the bottom.  By siloing these function into seperate layers, their behaviors and constraints can be designed seperately.}\label{fig:mechatronic_tech_stack}
\end{figure}

@@

The principal middleware requirement was an EtherCAT master, which itself depends on system-level networking. Linux satisfies the system-level networking requirement. EtherCAT masters are conventionally provided by PLC-based systems, like those from Beckhoff~\cite{BeckhoffTwinCAT}. However, they generally do not run on Linux. Additionally, their behavior is opque and does not lend itself to understanding the system. So, Ethercrab, a master written in the Rust programming language, was used~\cite{Ethercrab}. Ethercrab benefits from being open source, so motivated users can troubleshoot more easily. It also has thorough documentation and lots of examples (e.g.\ the control-layer application was based on \verb|ethercrab/examples/ek1100.rs|). Interfacing with Rust is a benefit due to its ease of use, quality tooling, performance, and memory safety~\cite{RustHomePage}.

Finally, the control layer had to use the API of the middleware layer to implement the business logic. One case of that logic was to have a controller to regulate the position of each cylinder. It had to be implemented in Rust to interface with the Ethercrab library. It also required some code to initialize the EtherCAT communication and to configure the Beckhoff terminals. There's also a user interface to interactively set the cylinder position. An interesting issue came up where the position sensor signal, mapping the $\qtyrange[range-units=single,range-phrase=..]{0}{10}{\volt}$ IO signal to the range 0..1~\verb|float|, but only reached the range 0..0.5~\verb|float|. This was because the step-down resistor was larger than designed since the right size wasn't available, so the maximum sensor reading was smaller than $\qty{10}{\volt}$. This unhandled gain is an undesired interface between the Control and IO layers since it demands the controller be written with knowledge of the minutia of how the IO was implemented. Further research will go toward how to accomodate this.

>>>>

In order to study mechatronics systems, it is necessary to design and develop physical research test beds. This is especially the case in mechatronic systems, as often they have component whose high-fidelity modeling is challenging and computationally expensive. When doing research with hardware systems, before a single research question can be asked of the hardware, researchers must assemble mechanical components, hook up power electronics, and route hoses for the hydraulic and pneumatic systems. Then the most complex of them all, the control system, must be added. Its tangle of spindly wires reaches all throughout the machine, meeting at the computer, the single most complex component of any fluid power system. The requirements placed on it are immense. In high-power systems, safety invariants must be held up perfectly lest disaster strike, and close monitoring is necessary due to the low stiffness of fluid systems compared to mechanical ones.

The tools used throughout history to create mechatronic systems have changed dramatically with developments in technology.  It will be enlightening to study it briefly.

Prior to the twentieth century, there weren't mechatronic systems as such, as the computer hadn't yet been invented, but there were controllers.  In the early eighteenth century, de Réamur created a device which regulated the temperature of a furnace which had a float on a vile of mercury that was connected through a linkage to a vent for the furnace~\cite{Bennett1996}.  When the temperature went up or down, the level of the mercury would change, causing the float to move and for the vent to open or close to encourage the furnace toward a more desired temperature.  Today we would understand this as simple proportional control, but enacted by fantastic mechanical means.

Another iconic design was the mechanical governor of the seventeenth century that regulated steam engine speeds by exploiting centrifugal forces to raise and lower masses to impede changes in engine speed.  This was another proportional controller.  It was described formally by Maxwell in 1868 where he wrote about its dynamics and stability, noting that the primary stability criterion is that all the roots of the systems dynamics must have negative real parts~\cite{Maxwell1868}.  Today we understand this as the Nyquist Criterion, but it was known 60 years before Nyquist wrote about it.

There were even proportional-derivative (PD) controllers developed in this era.  The US Navy developed the Whitehead Torpedo in 1873 which was able to regulate its depth based on a setting made by sailors before firing and then was able to dive to that depth and maintain level before hitting a target~\cite{Barber1874}.  It had a piston which was moved by hydrostatic pressures, and that piston connected to a control surface which set the torpedo's trajectory~\cite{Sears1873}.  It also had a pendulum which hung freely under gravity and a linkage connected it to another control surface on the torpedo.  Together, these made for a PD controller since the angle of the torpedo was related to the rate of change of depth.

These early controllers are incredibly fascinating designs, but their creativity was only exceeded by their complexity.  These designs coerced physical phenomena---gravity, pressure, temperature, rotation---into measurement signals that could be acted upon by a linkage-based controller.  Because the phenomena were myriad, there had to be a separate solution for every different type.  And the options for amplification were limited as well.  Mercury is not usually strong enough to open a furnace vent, so a lever will amplify the force.  But for a large furnace with a heavy door, there likely won't be a practical level.  What would ease the development of controllers would be a way to abstract phenomena into a single controllable signal.  Fortunately, the twentieth century saw the advent of exploitable electricity.

With electricity, measurement could be done by finding a way to convert a phenomenon to an electrical signal and that could be provided to a controller.  Sperry was able to electrically measure the state of the sea, wind, and a ship state to create an automatic ship steering controller~\cite{Bennett1996}.  Controllers also began to proliferate in industry where they manifested as boxes that could have measurements plugged in and a control signal would come out.  However, until Black developed feedback amplification~\cite{Black1934}, it was challenging to reliably amplify low-power control signals to a power level suitable to operate a control actuator.

One interesting case from this era of control was World War Two anti-aircraft guns.  These presented a difficult control problem.  Enemy aircraft had their position and velocity identified by radar (itself a difficult task at the time).  The controller had to anticipate where to aim the missile to intercept the target accounting for both the aircraft's and the missile's trajectories, all while tramming a heavy gun~\cite{Bennett1996}.  Initially, the measurement signal was provided by a worker shouting positions to the gun from where the radar was, but this proved too much latency for effective control.  The engineers working on the gun and those working on the radar needed to collaborate to be able to develop controller where the gun and radar would work as one automatic, closed-loop system.

In the post-war period, industrial controllers got much better thanks to the development of classical control theory in the 1930s and 40s.  However, electrical control proved to be challenging to implement.  Logic was typically implemented as huge relay circuits with dozens or hundreds of relays implementing logic~\cite{Parr1993}.  These relays were prone to failure and would cause downtime on factory lines when they needed to be replaced.  The fundamental issue was that up until this time, control programs were immutable once commissioned.  If there was a change needed, it would incur a substantial recommissioning process.  This would be solved by re-programmable controllers run on computers.

Industrial computers, or programmable logic controllers as they're called, were and are just regular computers, but the computer's package is made more robust to dust, temperature, vibrations, and other difficult environmental conditions found in industrial environments~\cite{Parr1993}.  They also have specialized programming languages that are meant to model to how control programs with relay logic were conceived.  But one major disadvantage of electronic controllers was still an issue.  The wires that ran to and from sensors and actuators made for a Cthulhu-esque tangle of wires and cables that were expensive and difficult to work on.  This was improved by the final tooling innovation discussed here, fieldbus networks.

Typically control and measurement signals consisted of $\qtyrange[range-units=single,range-phrase=..]{4}{20}{\milli\ampere}$ analog signals and $\qty{24}{\volt}$ digital signals.  This meant that every single sensor and actuator needed at least two wires, one for power and one for ground, going from it to the control box.  For large-scale controllers in factories, this was expensive and untenable to maintain~\cite{Thomesse2005}.  To solve this, practitioners developed fieldbus networks to handle their input and output.  The earliest was CAMAC which was used in the 1960s for nuclear instrumentation~\cite{Thomesse2005}.  Fieldbus networks provided benefits additional to wire cleanup.  They also enable distributing control information to many machines at once on a single network, monitoring the state of the factory floor from the design or business sections of the factory, and hierarchical control, such as described by the US National Bureau of Standards AMRF control hierarchy, where more abstract elements of an industrial plants command more concrete ones.  In particular, it describes a facility, consisting of production management and manufacturing engineering, commanding the shop, or factory floor, which in turn commands individual manufacturing cells, which then command the workstations within, and then at those workstations, operators command equipment~\cite{McLean1987}.  This all can be seen in figure~\ref{fig:amrf_control_hierarchy}.  The control tooling discussed in this thesis is concerned with the business at the Equipment and Workstation level.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{assets/amrf_control_hierarchy}
  \caption{The AMRF control hierarchy, boxed, abstracts different levels of behavior within a factory into discrete functions with well-defined interfaces between them.  Taken from~\cite{McLean1987}}\label{fig:amrf_control_hierarchy}
\end{figure}

This history has been brief, it doesn't go into the more modern tooling in robotics and its dependencies on signal libraries like ROS~\cite{ROS2}.  But what it has shown is the importance of abstraction.  Early mechanical controllers were challenging because every different physical phenomenon had to be separately and mechanically integrated into the controller.  Electricity simplified that.  Complex electrical controllers proved expensive and difficult to maintain and modify until computers moved logic to the re-programmable realm.  Finally, IO saw direct signaling abstracted to bus communication reducing massive tangles of wire to only a single cable.

This paper builds on this history to improve the process of commissioning control systems for mechatronic systems by reasoning about how to abstract the construction of behavior and components in computer control systems into layers, synthesizing a novel {\it mechatronic tech stack\/} theory, not unlike the AMRF control hierarchy. It will demonstrate in three case studies how this theory can be applied in mechatronic systems. The tools used in these cases will be categorized into layers of the mechatronic tech stack.  The analysis shows that, while different, these cases all fit into the same structure.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/osi_model}
  \caption{The OSI model understands network communication to be a layer cake of abstractions over the physical communication media.  An individual tool may implement the behaviors and interfaces of of one or multiple layers.  The layering enables design freedom in how a network is implemented.  Communication over HTTP usually goes over a TCP transport through an IP network and to other destinations over WiFi or Ethernet, but Lindgren et al.\ show that it's possible to send IP signals over CAN~\cite{Lindgren2008}, and Ethernet over EtherCAT is possible as well~\cite[§1.9.3]{EtherCATFieldbus}.  High level protocols can also be replaced to obtain a lower level of abstraction for performance or design flexibility, as was done for the video game \emph{DOOM} in 1993, as well as for contemporary game engines like Godot or Unity~\cite{DOOM1993,GodotMultiplayer,UnityTransport}.}\label{fig:osi_model}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/printf_syntax}
  \cprotect\caption{The above printf example with syntax labeled according to Kernighan {\&} Ritchie C~\cite{Kernighan1978}.  The whole line is a postfix expression and the \verb|printf| identifier is also a postfix expression and a primary expression on its own.  The tokens between parentheses are an argument expression list, and within that, the tokens between commas are at once many types of expressions that enable writing non-trivial code, but terminating as a primary expression, and then here as a string literal and an identifier.}\label{fig:printf_syntax}
\end{figure}

The semantics of \verb|printf| are the expected behavior of the function, which can be gleaned from the documentation. It will say that \verb|printf| means to convert and write output to the console under the control of a format string~\cite{Kernighan1978}.  It will also say that the format string is the first expression in the argument expression list and that it is of type \verb|char*|.  So \verb|"I have %u cows\n"| is that format string, and the documentation will state that the second expression in the argument expression list, here \verb|num_cows| will be formatted as a string and written to the output where conversion character, here \verb|%u| is.  The logic of how the conversion characters and string formatting work are based on the format string's own morphology, syntax, and semantics, the study of which is left as an exercise to the reader.

@@

Syntax is the structure of an idea. It is a finer comb to discriminate layers. Software generally is built as a linear sequence of instructions with branches and loops. This describes the control engineer's perspective on their own control program, but supporting software like libraries are more abstract. They're better thought of as a dependency. A control program may depend on libraries A, B, and C, and those in turn depend on libraries D, E, F, and G. These \textit{middlewares}, named so as they exist in the middle between the control software and the computer, are better understood as a network of the programs themselves and the dependency connections between them. For hardware, its structure is its assemblage. The computer components are connected to the motherboard, sensors and transducers are connected to a CAN bus, the temperature sensors message over I2C, etc. Like middleware, hardware is a network of components and the connections between them, but the type of connection is also important. A system with components on a CAN bus is going to be different than one with a USB bus, and different still from a system where all components communicate over a serial connection.

@@

This is enough to propose a mechatronic tech stack. The mechatronic stack as outlined sits on top of signaling with sensors and transducers. Those are interfaced by IO systems that connect to computer systems which run middleware upon which a control program depends. The design is so that functionality in the control layer should be preserved if components in the lower layers are changed, perhaps with some implementation effort, as was the case for IP over CAN.  This is illustrated as a stack diagram in figure~\ref{fig:mechatronic_tech_stack}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/mechatronic_tech_stack}
  \caption{The mechatronic tech stack places high-level business logic at the top and low-level interfacing at the bottom.  By siloing these function into separate layers, their behaviors and constraints can be designed separately.}\label{fig:mechatronic_tech_stack}
\end{figure}

@@

The principal middleware requirement was an EtherCAT master, which itself depends on system-level networking. Linux satisfies the system-level networking requirement. EtherCAT masters are conventionally provided by PLC-based systems, like those from Beckhoff~\cite{BeckhoffTwinCAT}. However, they generally do not run on Linux. Additionally, their behavior is opaque and does not lend itself to understanding the system. So, Ethercrab, a master written in the Rust programming language, was used~\cite{Ethercrab}. Ethercrab benefits from being open source, so motivated users can troubleshoot more easily. It also has thorough documentation and lots of examples (e.g.\ the control-layer application was based on \verb|ethercrab/examples/ek1100.rs|). Interfacing with Rust is a benefit due to its ease of use, quality tooling, performance, and memory safety~\cite{RustHomePage}.

Finally, the control layer had to use the API of the middleware layer to implement the business logic. One case of that logic was to have a controller to regulate the position of each cylinder. It had to be implemented in Rust to interface with the Ethercrab library. It also required some code to initialize the EtherCAT communication and to configure the Beckhoff terminals. There's also a user interface to interactively set the cylinder position. An interesting issue came up where the position sensor signal, mapping the $\qtyrange[range-units=single,range-phrase=..]{0}{10}{\volt}$ IO signal to the range 0..1~\verb|float|, but only reached the range 0..0.5~\verb|float|. This was because the step-down resistor was larger than designed since the right size wasn't available, so the maximum sensor reading was smaller than $\qty{10}{\volt}$. This unhandled gain is an undesired interface between the Control and IO layers since it demands the controller be written with knowledge of the minutia of how the IO was implemented. Further research will go toward how to accommodate this.
